---
title: "Obtain variance for difference-in-difference from a Poisson log-linear model using the Delta Method"
author: "Jade Benjamin-Chung"
date: "12/5/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(msm)
library(dplyr)
source(here::here("0-config.R"))

data_zip = readRDS(paste0(data_dir, "ceip-flu-data-age-sex-zip.RDS"))

data = data_zip$All$all

d1415 = data[data$seas>=1112 & data$seas!="1516" & data$seas!="1617" & data$seas!="1718",]
d1415$time=ifelse(d1415$seas==1415,1,0)

```

# Objective

When using count data to obtain estimates of incidence rates, log-linear models are commonly since a population-offset can be conveniently included in the model, so individual-level data is not required. This document uses the Delta Method to obtain the standard error for a relative scale difference-in-difference parameter using the coefficients from a log-linear model, fit as follows: 

$\ln\bigg(\frac{E[Y|X, T]}{PT} \bigg) = \beta_0 + \beta_1X + \beta_2T + \beta_3 X\cdot T$

- $Y$: count outcome
- $PT$: person-time used to calculate rate
- $X$: indicator for treatment
- $T$: indicator for time

The parameter of interest is:
$$\text{RDID} = 1 - \bigg( \frac{E[Y|X=1, T=1]}{E[Y|X=0, T=1] + (E[Y|X=1, T=0] - E[Y|X=1, T=0])}  \bigg) \times 100$$
  
Using the coefficients from the model above, the transformation function is:

$$ F(\beta) = \bigg(1 - \frac{e^{\beta_0 + \beta_1 + \beta_2 + \beta_3 }}{ e^{\beta_0 + \beta_2} + e^{\beta_0 + \beta_1} - e^{\beta_0}}\bigg) \times 100 $$


# Calculate relative scale difference-in-difference by hand
```{r hand}
# Pre-program numerator and denominator
data %>% group_by(dist) %>% 
  filter(seas>=1112 & seas<1415) %>% 
  group_by(seas, dist) %>%
  summarise(flucases = sum(flucases), pop = sum(N)) %>%
  group_by(dist) %>%
  summarise(flucases = mean(flucases), pop = mean(pop)) 

# 2014-15 numerator and denominator
data %>% filter(seas==1415) %>% 
  group_by(dist) %>%
  summarise(flucases = sum(flucases), pop = sum(N)) 

# rr
(158/447832) / ((113/255318) + (117/447832) - (56/255318))

# 1-rr x 100%
(1-(158/447832) / ((113/255318) + (117/447832) - (56/255318)))*100
```

# Obtain ratio of incidence ratios using Poisson model

```{r poisson}
glm.fit=glm(flucases ~ tr*time,offset=logN,data=d1415,
             family=poisson(link=log))
```

# Use Delta Method to get variance for difference-in-difference

## Step 1: Get partial derivatives of $F(\beta)$ with respect to each parameter 

$$\frac{dF}{d\beta_0} = 0$$

$$\frac{dF}{d\beta_1} = \frac{(e^{\beta_2 }-1)e^{\beta_1 + \beta_2 + \beta_3}}{(e^{\beta_1} +e^{\beta_2}-1)^2} $$

$$\frac{dF}{d\beta_2} = \frac{(e^{\beta_1}-1)e^{\beta_1+\beta_2+\beta_3}}{(e^{\beta_1}+e^{\beta_2}-1)^2} $$

$$\frac{dF}{d\beta_3} = \frac{e^{\beta_1 + \beta_2 + \beta_3 }}{e^{\beta_1}+e^{\beta_2}-1} $$


```{r betas}
# Gradient 
b0 = glm.fit$coefficients[1]
b1 = glm.fit$coefficients[2]
b2 = glm.fit$coefficients[3]
b3 = glm.fit$coefficients[4]

dfdb0 = 0
dfdb1 = ((exp(b2)-1)*exp(b1+b2+b3)) / ((exp(b1) + exp(b2)-1)^2)
dfdb2 = ((exp(b1)-1)*exp(b1+b2+b3)) / ((exp(b1) + exp(b2)-1)^2)
dfdb3 = (exp(b1 + b2 + b3))/(exp(b1) + exp(b2) - 1)

grad <- c(dfdb0, dfdb1, dfdb2, dfdb3)
names(grad) <- c("dfdb0", "dfdb1", "dfdb2", "dfdb3")
grad
```


## Step 2: Get the covariance variance matrix from regression output 

```{r vcov}
vb <- vcov(glm.fit)
vb
```


## Step 3: Calculate the variance 
$Var(F(\beta)) = J V J^T$, where $J$ is the Jacobian or gradient (derivative of $F(\beta)$ with respect to each parameter) and $V$ is the variance covariance matrix. 

```{r vcovdelta}
vF <- t(grad) %*% vb %*% grad
vF

seF <- sqrt(vF)
seF
```

## Obtain relative scale difference-in-difference and 95% CI from log-linear model

```{r makefun}
get_rdid = function(fit){
  
# Gradient 
b0 = fit$coefficients[1]
b1 = fit$coefficients[2]
b2 = fit$coefficients[3]
b3 = fit$coefficients[4]

dfdb0 = 0
dfdb1 = ((exp(b2)-1)*exp(b1+b2+b3)) / ((exp(b1) + exp(b2)-1)^2)
dfdb2 = ((exp(b1)-1)*exp(b1+b2+b3)) / ((exp(b1) + exp(b2)-1)^2)
dfdb3 = (exp(b1 + b2 + b3))/(exp(b1) + exp(b2) - 1)

grad <- c(dfdb0, dfdb1, dfdb2, dfdb3)
names(grad) <- c("dfdb0", "dfdb1", "dfdb2", "dfdb3")
grad

# Estimate relative scale DID
rr = (exp(b0+b1+b2+b3))/( exp(b0+b2) + exp(b0+b1) - exp(b0))
rdid = (1-rr) * 100 
print(paste("RR adjusted for pre-intervention = ", 
            sprintf("%0.02f",rr), sep = " "))
print(paste("Relative scale DID = ", sprintf("%0.0f",rdid), sep = " "))

# Variance-covariance matrix
vb <- vcov(fit)
vb

# Variance of DID
vF <- t(grad) %*% vb %*% grad
vF

se <- sqrt(vF)

# 95% CI 
lb = log(rr) - (qnorm(0.975)*sqrt(vF))
ub = log(rr) + (qnorm(0.975)*sqrt(vF))

rdid = (1 - rr)*100
lb = (1 - exp(lb))*100
ub = (1 - exp(ub))*100

return(list = c(rdid = rdid, se = se, lb = ub, ub = lb))

}

get_rdid(glm.fit)

```

## Use deltamethod package to obtain relative scale difference-in-difference and 95% CI from log-linear model

```{r pkg}
g = as.formula(~ (exp(x1+x2+x3+x4) /(exp(x1+x2) + exp(x1+x3) - exp(x1))))
deltamethod(g = g, mean = coef(glm.fit), cov = vcov(glm.fit))

```

